---
title: "Section C3" 
format: html 
editor: visual 
---

# C.3 Classification Analysis

## Four most significant variables

```{r}
nhanes <- read.csv("NHANES3_419.csv", header = TRUE, sep = ",")  
#Omit height 
data <- nhanes[,-4]
```

## Classification Analysis

```{r}
lin.class <- function(Y,group){   
  # Install MASS package if not already installed   
  if (!require("MASS")) install.packages("MASS") 
  library(MASS)   
  Y <- data.matrix(Y)   
  group <- as.factor(group)   
  p <- ncol(Y)   
  m1 <- manova(Y ~ group)   
  nu.e <- m1$df   
  nu.h <- m1$rank-1   
  Sp <- summary(m1)$SS$Residual/(nu.e)   
  cio <- 1:m1$rank   
  c.mat <- matrix(nrow=m1$rank,ncol=p,0)   
  for (i in 1:m1$rank) {     
    cio[i] <- -.5*t(lda(Y,group)$means[i,])%*%solve(Sp)%*%       
      lda(Y,group)$means[i,]     
    c.mat[i,] <- t(lda(Y,group)$means[i,])%*%solve(Sp)   
  }   
  return(list("coefs"=c.mat,"c.0"=cio)) }  

lin.class(data[,-1], data[,1])
```

Given:

y1 = Age (Variable 2)

y2 = Body Weight (Variable 3)

y3 = Average Diastolic BP (Variable 5)

y4 = Serum Cholesterol (Variable 6)

Then:

L1 = 0.292y1 + 0.112y2 + 0.679y3 + 0.079y4 - 45.528

L2 = 0.356y1 + 0.129y2 + 0.781y3 + 0.073y4 - 57.051

L3 = 0.446y1 + 0.135y2 + 0.838y3 + 0.063y4 - 65.632

Assign to G1, G2, or G3 depending on if L1, L2, or L3 respectively yields the greatest value for a given observation.

## Apply Classification to Observation 1

Observation 1, eliminating height:

nhanes(1)'= (63 141.4 64 20)

```{r}
L1 <-  0.2916651%*%(63) + 0.1120130%*%(141.4) + 0.6792672%*%(64) + 0.07939680%*%(220) - 45.52791  
L2 <-  0.3555073%*%(63) + 0.1287796%*%(141.4) + 0.7809797%*%(64) + 0.07280753%*%(220) - 57.05103  
L3 <-  0.4457149%*%(63) + 0.1350911%*%(141.4) + 0.8384779%*%(64) + 0.0634611%*%(220) - 65.63240  

L1 
L2 
L3
```

For a single observation, the linear classification function with the greatest returned value is considered to be assigned to that group. For the first observation, the first linear classification function returned the greatest value of 49.626, indicating that the observation be correctly classified as part of Group 1.

This is very similar to the values of other functions L2 and L3 (49.556 and 49.174 respectively), which could indicate a large amount of noise in the data. Non-linear classification functions might explain separation of groups better if further analysis were to be conducted.

## Confusion Matrix

```{r}
rates <- function(data,group,method="l") {   
  if (!require("MASS")) install.packages("MASS")   
  library(MASS)   
  data <- as.matrix(data)   
  group <- as.matrix(group)   
  da.obj <- lda(data,group)   
  if (method=="q") {     
    da.obj <- qda(data,group)     
    method <- "QDA"   }   
  tab <- table(original=group,predicted=predict(da.obj)$class)   
  if (method=="l") method <- "LDA"   
  cor.rate <- sum(predict(da.obj)$class==group)/nrow(data)   
  er.rate <- 1-cor.rate   
  return(list("Correct Class Rate"=cor.rate,"Error Rate"=er.rate,               
              "Method"=method,"Confusion Matrix"=tab
              )) 
  }  

rates(data[,-1], data[,1])
```

ACCR = 63.64%

AER = 36.36%

The linear classification functions are classifying the majority of observations correctly (64%), but there is some classification error (36%). This supports earlier conclusions of high noise within the data leading to some incorrectly classified observations.
